---
title: "Lab 7 Assignment: Group 29"
format:
  html:
    embed-resources: true
editor: visual
---

## Load libraries

```{r}
library(tidyverse)
library(broom)

```

## Load data

```{r}

raw_dir <- "data/_raw/"
data_file <- "gravier.RData"
data_loc <- "https://github.com/ramhiser/datamicroarray/raw/master/data/"

if( !dir.exists(raw_dir) ){
  dir.create(path = raw_dir)
}
if( !file.exists(str_c(raw_dir, data_file)) ){
  download.file(
    url = str_c(data_loc, data_file),
    destfile = str_c(raw_dir, data_file))
}
load(file = str_c(raw_dir, data_file))

```

The data frame is then converted to a tibble and early_metastasis is added as a column in the beginning.

```{r}

gravier <- gravier |>
  bind_cols() |>
  as_tibble() |>
  mutate(y = case_when(y == "poor" ~ "poor",
                       y == "good" ~ "good")) |> 
  relocate(early_metastasis = y) |>
  select(1:101)
```

```{r}
gravier_pca <- gravier|>
  select(2:6)
```

```{r}
pca_fit <- gravier_pca %>% 
  select(where(is.numeric)) %>% # retain only numeric columns
  prcomp(scale = TRUE) # do PCA on scaled data
```

```{r}
pca_fit %>%
  tidy(matrix = "rotation")|>
  group_by(column) |> 
  slice(1:2) |> 
  ungroup()
```

```{r}
library(ggplot2)
arrow_style <- arrow(
  angle = 30, ends = "first", type = "closed", length = grid::unit(5, "pt")
)

# plot rotation matrix
pca_plot <- pca_fit %>%
  tidy(matrix = "rotation") %>%
  pivot_wider(names_from = "PC", names_prefix = "PC", values_from = "value") %>%
  ggplot(aes(PC1, PC2)) +
  geom_segment(xend = 0, yend = 0, arrow = arrow_style) +
  geom_text(
    aes(label = column),
    hjust = 1, nudge_x = -0.01, 
    color = "#FF0000"
  ) +
  xlim(-0.75, 0.01) + ylim(-0.75, 0.5) +
  coord_fixed(ratio = 1) +
  theme_minimal(12)
```

Here you see a PCA plot showing how the selected genes contribute to the first two principal components. Each arrow represents a gene (see labels in red). The direction and length indicates how strongly it correlates with the respective PC.

```{r}
print(pca_plot)
```
